---
title: "**Maternal dietary behaviors vs Asthma in children**"
output: html_document
---

##### **Group members:** Juhi Mattekatt (jm5916) and Nichcha Subdee (ns3923)

---

#### **Final project option 3 (Group assignment)**

**Dataset:** Exposome

**Type:** Hypotheses generation 

**Research question:** What maternal dietary behaviors contribute to asthma incidence in offspring?

---

#### **Background** 

Asthma is one of the most common chronic diseases in children worldwide. It affected approximately 4,313 per 100,000 children under 20 years old in 2021 and accounts for more than 1% of global disability-adjusted life years (DALYs) (Kim et al., 2025). Although the global prevalence has declined by about 30% since 1990, asthma is expected to remain a major health burden through 2050, with the highest rates seen in children aged 5–9 years. Early-life and prenatal factors are known to play an important role in asthma development, and maternal health behaviors during pregnancy are one key area of interest. Maternal diet is a modifiable factor that may affect fetal immune development and airway formation, yet it remains unclear which specific dietary behaviors are most strongly linked to childhood asthma. 

Therefore, this study will apply machine learning approaches to explore the relationship between maternal dietary behaviors during pregnancy and asthma in children aged 6–11 years.

---

#### **Objective**

* To explore the relationship between maternal dietary behaviors during pregnancy and asthma in children aged 6–11 years.

* To compare the performance of different machine learning algorithms in detecting the relationship between maternal dietary behaviors during pregnancy and childhood asthma risk.

* To determine  maternal dietary behaviors during pregnancy that have the greatest risk of leading to asthma in offspring for future research.

---

#### **Methodology**

This study uses a real world exposome dataset based on the HELIX project database. The HELIX project is a multi-center European birth cohort integrating six longitudinal cohorts from France, Greece, Lithuania, Norway, Spain, and the UK, with children born between 2003 and 2009. 

A cohort of 1,301 mother–child pairs was followed up at ages 6–12 years, with health outcomes including doctor-diagnosed asthma, dietary exposures, and sociodemographic characteristics recorded across all centers (Agier et al., 2019).

**Outcome:** `hs_asthma` -- Doctor diagnosed asthma (ever)

**Example Dietary features:** 8 variables for maternal dietary behaviors during pregnancy were found

* `h_cereal_preg_Ter` -- cereal consumption during pregnancy (times/week)

* `h_dairy_preg_Ter` -- dairy consumption during pregnancy (times/week)

* `h_fastfood_preg_Ter`	-- fast food consumption during pregnancy (times/week)

* `h_fish_preg_Ter` -- fish consumption during pregnancy (times/week)

* `h_fruit_preg_Ter` -- fruit consumption during pregnancy (times/week)

* `h_legume_preg_Ter` -- legume consumption during pregnancy (times/week)

* `h_meat_preg_Ter` -- meat consumption during pregnancy (times/week)

* `h_veg_preg_Ter` -- vegetables consumption during pregnancy (times/week)

_Note: All dietary variables are tertile-coded (1 = low, 2 = medium, 3 = high)_

**Study Framework:**

1. **Data preparation & cleaning**

2. **Unsupervised learning** (K-Means Clustering) 

3. **Data partition** (70/30 stratified train/test split)

4. **Supervised learning with hyperparameter tuning** (10-fold cross-validation)

* Logistic Regression
* Elastic Net 
* Random Forest 

5. **Model evaluation** (AUC-ROC, Sensitivity, Specificity on test set)

6. **Interpretation** 

#### **Pipeline**

```{r library, include = FALSE}
library(tidyverse)
library(caret)
library(glmnet)
library(cluster)
library(factoextra)
library(e1071)
library(randomForest)
library(pROC)
```

**Step 1: Data Preparation & Cleaning ** 

```{r import, message=FALSE, include = FALSE}
# Load data
load("exposome.RData")

# Merge all data frames
studydata = merge(exposome, phenotype, by = "ID") |> 
  merge(covariates, by = "ID")

# Remove ID
studydata$ID = NULL

# Select variables
diet_vars = c("h_cereal_preg_Ter", "h_dairy_preg_Ter", "h_fastfood_preg_Ter",
               "h_fish_preg_Ter", "h_fruit_preg_Ter", "h_legume_preg_Ter",
               "h_meat_preg_Ter", "h_veg_preg_Ter")

df = studydata[, c("hs_asthma", diet_vars)]

# Recode outcome as factor
df$hs_asthma = factor(df$hs_asthma, levels = c(0,1), labels = c("No", "Yes"))

# Ensure dietary variables are ordered factors
df[, diet_vars] = lapply(df[, diet_vars], function(x) factor(x, ordered = TRUE))

# Complete-case analysis
df = na.omit(df)
```

```{r data_summary}
# Final dataset dimensions and outcome distribution
dim(df)
table(df$hs_asthma)
```

The final analytic dataset contained **1,301** mother–child pairs and **9** variables (1 outcome, 8 maternal dietary exposures). The outcome variable `hs_asthma` showed class imbalance (1,159 children without asthma vs. 142 children with a doctor-diagnosed asthma). This suggests the need for resampling during model training.

**Step 2: Unsupervised Learning — K-Means Clustering**

The optimal number of clusters (k) was determined using two complementary approaches: the gap statistic and the elbow method.

```{r kmeans_gap, echo=FALSE, warning=FALSE}
set.seed(123)

diet_matrix = df[, diet_vars] %>% 
  mutate(across(everything(), as.numeric)) %>% 
  scale()

#Gap-statistic
gap_stat = clusGap(diet_matrix, FUN = kmeans, nstart = 25, K.max = 8, B = 50)
fviz_gap_stat(gap_stat)
print(gap_stat, method = "firstmax")

#Elbow method
fviz_nbclust(diet_matrix, kmeans, method = "wss") +
  labs(title = "Elbow Method for Optimal k",
       x     = "Number of Clusters (k)",
       y     = "Total Within-Cluster Sum of Squares")
```

The gap statistic recommended k = 8 as the optimal solution (gap = 0.549, SE = 0.005). However, since the analysis included exactly 8 dietary variables, using k = 8 would essentially assign each variable its own cluster, which limits meaningful interpretation. In addition, the elbow plot showed the largest drop in within-cluster variation from k = 1 to k = 3. Based on both methods, **k = 3 was chosen as the final number of clusters to represent three distinct maternal dietary patterns.**

```{r kmeans_fit, echo=FALSE}
set.seed(123)

km = diet_matrix %>%
  kmeans(centers = 3, nstart = 25)

fviz_cluster(km, data = diet_matrix)

cluster_profile = cbind(
  df[, diet_vars] %>% mutate(across(everything(), as.numeric)),
  diet_cluster = km$cluster
)

cluster_profile %>%
  group_by(diet_cluster) %>%
  summarise_all(mean) %>%
  knitr::kable(digits = 2,
               caption = "Table 1. Mean Dietary Intake by Cluster")

df$diet_cluster = factor(km$cluster)
table(df$diet_cluster)
```

Three distinct maternal dietary patterns were identified using k-means clustering (k = 3), based on mean tertile scores across eight dietary food groups during pregnancy.

* **Cluster 1: High Legume/Dairy** (n = 657)  

The highest legume (mean = 2.95) and dairy (mean = 2.69) intake among all clusters, suggesting a diet rich in plant-based protein and animal-sourced nutrients. Fast food consumption was also elevated (mean = 2.64).

* **Cluster 2: High Cereal/Meat** (n = 268)  

The highest cereal (mean = 2.74), meat (mean = 2.35), and fish intake (mean = 2.29), reflecting a more Western-style dietary pattern with higher processed grain and animal protein consumption.

* **Cluster 3: High Vegetable/Fruit** (n = 376)  

The highest vegetable (mean = 2.57) and fruit (mean = 2.47) intake, suggesting a diet with greater emphasis on plant-based foods. 

_Note:_ From **table 1**, fast food consumption was similarly elevated in Cluster 1 (2.64) and Cluster 3 (2.63), and significantly lower in Cluster 2 (1.71). This pattern may partly reflect the mixed nutritional composition of fast food items common, which usually include dairy, vegetables, and legumes. This probably makes elevated fast food consumption behaviorally compatible with otherwise distinct dietary profiles.

**Step 3: Data Partition**

The dataset was split into 70% training and 30% test sets using stratified partitioning to preserve the class distribution of `hs_asthma`.

```{r partition, include=TRUE, echo=FALSE}
set.seed(123)

train_index = createDataPartition(df$hs_asthma, p = 0.7, list = FALSE)
train_data  = df[train_index, ]
test_data   = df[-train_index, ]

# Show sizes
cat("Training set:", nrow(train_data), "observations\n")
cat("Test set:    ", nrow(test_data), "observations\n")

```

The dataset was randomly split into a training set (n = 912) and a 
test set (n = 389).

**Step 4: Supervised Learning (Logistic Regression, Elastic Net, and Random Forest)**

Up-sampling was applied to address class imbalance issues.

10-fold cross validation was used as it provides the most robust estimates for model performance, our sample size wasn't too large so the computational cost was not too high. 

The models were chosen as they are more interpretable than flexible, which is preferred when doing hypothesis-generation.

_Model 1:_ Logistic regression was selected as it is useful in hypothesis generation when trying to find statistically significant associations, it is also the standard baseline. 

There are no hyperparameters to tune here.

```{r logistic_train, include=TRUE, warning=FALSE}
set.seed(123)

# 10-fold CV with upsampling for class imbalance
control_settings = trainControl(
  method          = "cv",
  number          = 10,
  sampling        = "up",
  classProbs      = TRUE,
  summaryFunction = twoClassSummary
)

l_model<- train(
  hs_asthma ~ .,
  data = train_data,
  method = "glm", 
  trControl = control_settings,
  preProcess = c("center","scale")
)
```

_Model 2:_ Elastic Net was selected to handle correlation among dietary variables. 

```{r en_train, include=TRUE}
set.seed(123)

# Broad lambda grid
lambda   = 10^seq(-3, 1, length = 100)
en_grid  = expand.grid(alpha = seq(0, 1, by = 0.1), lambda = lambda)

en_model = train(
  hs_asthma ~.,
  data      = train_data,
  method    = "glmnet",
  preProc   = c("center", "scale"),
  trControl = control_settings,
  tuneGrid  = en_grid,
  metric    = "ROC"
)
```

_Model 3:_ Random Forest was selected as it can handle non-linear relationships and interaction well.

```{r rf_train, include=TRUE}
set.seed(123)

#Broad mtry
num_predictors <- ncol(train_data) - 1
mtry.vals <- c(num_predictors, sqrt(num_predictors), 0.5 * num_predictors)
mtry.grid <- expand.grid(.mtry = round(mtry.vals))

rf_model <- train(
  hs_asthma ~ ., 
  data = train_data, 
  method = "rf", 
  metric = "ROC", 
  tuneGrid = mtry.grid, 
  trControl = control_settings, 
  ntree = 100,
  importance = TRUE
)
```

**Step 5: Model Comparison** 

```{r mod_comp, echo=FALSE}
model_comparison <- resamples(list(
  Logistic = l_model,
  Elastic_Net = en_model,
  Random_Forest = rf_model
))

mc <- summary(model_comparison)
summary <- mc$statistics$ROC[, "Mean"]

summary

bwplot(model_comparison, metric = "ROC")
```

The elastic net model had the largest mean ROC of `r round(summary["Elastic_Net"], 2)`. We will apply our elastic net model to our test set. 

**Step 6: Apply Model**

```{r app_mod, echo=FALSE}
final <- predict(en_model, test_data, type = "prob")
test_roc <- roc(test_data$hs_asthma, final$Yes)
plot(test_roc, main = "Final Model ROC Curve on Test Data")

sen_spec <- predict(en_model, newdata = test_data)
confusionMatrix(sen_spec, test_data$hs_asthma, positive = "Yes")
```

This model has weak sensitivity and weak specificity. It is also only accurate in its predictions about half the time. This suggests that the features included in this model don’t contain enough predictive power to predict risk of asthma.

**Step 7: Variable Importance **

```{r var_imp, echo=FALSE}
varImp(en_model)
plot(varImp(en_model))
```

Cohort of inclusion was the most influential feature in this model. The most important dietary predictors were meat and fish consumption during pregnancy.

#### **Interpretation**
Results from our analytic pipeline indicate that the selected dietary and demographic features did not have enough predictive power to accurately predict asthma cases. However, while our model was not highly accurate, our findings suggest that meat consumption and fish consumption during pregnancy might play a role in asthma incidence. Future research should consider these exposures.  

Our findings should not be used in clinical or real-world settings. No causal inference can be made from our findings due to this study being hypothesis generating in nature. We can only recommend that research into maternal diet and asthma be informed by our findings. 


#### **Limitations**

A major limitation of the research is that the dietary variables included were tertile-coded, which may obscure true variation in intake and limit precision of the dietary patterns identified. By assuming constant risk within each tertile, we likley lost statistical power. This may explain the lack of predictive power that we found. Additionally, the HELIX cohort is limited to six European countries, which may limit generalizability to populations outside Europe. We also used up-sampling to handle the unbalanced asthma cases and controls. This may have led to overfitting which would further reduce the generalizability of our results

#### **Ethical Considerations**

Dietary patterns are shaped by socioeconomic and cultural factors. Results should not be used to stigmatize individuals with limited food access, and any future research should account for structural barriers to healthy eating. Moreover, our dataset relies on European data and best informs research done on similar populations. For instance, research into meat and fish consumption as a cause for asthma may not be beneficial in areas with greater vegetarian populations. 

---

#### **References**

1. Kim, Tae Hyeon et al. “Global burden of asthma among children and adolescents with projections to 2050: a comprehensive review and forecasted modeling study.” _Clinical and experimental pediatrics_ vol. 68,5 (2025): 329-343. doi:10.3345/cep.2025.00423

2. Agier, Lydiane, et al. "Early-life exposome and lung function in children in Europe: an analysis of data from the longitudinal, population-based HELIX cohort." _The Lancet Planetary Health_ 3.2 (2019): e81-e92.

---

#### **AI tools**

Claude.ai was used in this assignment to assist with debugging R code syntax errors and editing text for clarity and conciseness.
